New approaches for boosting to uniformity

### 2014-10-04 assets/paperdraft/parts/dataset.tex

Figure~\ref{fig:dalitz} shows the Dalitz-plot distributions for signal and background events.  These samples are split into training and testing samples and then various BDTs are trained.  Figure~\ref{fig:dalitz_results} shows the efficiency obtained for each classifier {\em vs} distance from the a corner of the Dalitz-plot.  The AdaBoost algorithm, as expected, produces a much lower efficiency in the interesting corner regions.  The kNNAdaBoost algorithm does not improve upon the AdaBoost result much.  This is likely due to the fact that while kNNAdaBoost uses non-local kNN information, it does not utilize global information.
The UGBFL (binned and unbinned kNN) and uBoost algorithms each produce an efficiency which is statistically consistent with uniform across the Dalitz plot.

\begin{figure}[] 
  \centering 
  \includegraphics[width=0.49\textwidth]{ROC_DP.pdf}
  \includegraphics[width=0.49\textwidth]{ROC_DP_Alpha.pdf}
  \caption{\label{fig:dalitz_rocs} Efficiency {\em vs} distance to a corner of the Dalitz-plot.  An arbitrary working point of 50\% integrated efficiency is displayed.}
\end{figure}

@@

ADD DISCUSSION ON ROC CURVE HERE ...

\begin{figure}[] 
  \centering 
  \includegraphics[width=0.49\textwidth]{MD_eff.pdf}
  \caption{\label{fig:dalitz_results} Background mis-identification {\em vs} $D$ candidate mass for (black circles) AdaBoost, (cyan up triangles) kNNAdaBoost, (red filled squares) UGBFL(bin), (red open squares) UGBFL(knn), and (blue down triangles) uBoost.  An arbitrary working point of 10\% background mis-identification in the training region $1.75 < m(D) < 1.85$~GeV is displayed.}
\end{figure}

>>>>

Figure~\ref{fig:dalitz} shows the Dalitz-plot distributions for signal and background events.  These samples are split into training and testing samples and then various BDTs are trained.  For the BDTs designed to produce uniform selections, the $\vec{y}$ variates are the Dalitz masses with the choice of uniform selection efficiency on signal candidates in the Dalitz-plot.  
Figure~\ref{fig:dalitz_rocs} shows the ROC curves obtained for the various classifiers studied in this paper.  For the UGBFL algorithms, there is a choice to be made for the value $\alpha$ which defines the relative weight of the flatness loss {\em vs} AdaLoss.  As expected, increasing $\alpha$, which increases the weight of AdaLoss, drives the ROC curve to be similar to AdaBoost.  Analysts will need to choose how much ROC performance to sacrifice to gain uniformity in the selection efficiency.  In general, the ROC curves for the uniform-driven BDTs are not too different from AdaBoost.

\begin{figure}[] 
  \centering 
  \includegraphics[width=0.49\textwidth]{ROC_DP.pdf}
  \includegraphics[width=0.49\textwidth]{ROC_DP_Alpha.pdf}
  \caption{\label{fig:dalitz_rocs} (left) ROC curves for classifier algorithms studied in this paper.  (right) ROC curves for UGBFL(bin) for differnet values of $\alpha$.}
\end{figure}

Figure~\ref{fig:dalitz_results} shows the efficiency obtained for each classifier {\em vs} distance from the a corner of the Dalitz-plot\footnote{The Dalitz-plot is essentially a triangle with three corners. Our defnition of this distance is ${\rm MIN}\left[(m(D_s)-m(\pi))^2 - m_{ij}^2\right]$, where $ij$ is 12, 13 and 23.}.  The AdaBoost algorithm, as expected, produces a much lower efficiency in the interesting corner regions.  The kNNAdaBoost algorithm does not improve upon the AdaBoost result much.  This is likely due to the fact that while kNNAdaBoost uses non-local kNN information, it does not utilize global information.
The UGBFL (binned and unbinned kNN) and uBoost algorithms each produce an efficiency which is statistically consistent with uniform across the Dalitz plot.  
As stated above, the analyst is free to optimize the choice of $\alpha$ for UGBFL by defining a metric that involves signal efficiency, background rejection and uniformity, {\em e.g.}, using uniformity metrics discussed in detail in Appendix~A.

@@

As a separate study using the same data samples, consider the case where one has simulated signal events and uses data from a {\em nearby} region, a so-called sideband, for background.  This is a common situation in particle-physics analyses.  Figure~\ref{fig:md_results} shows the training samples used.  A major problem can arise in these situations as typically input variates to the BDT are correlated with the parent particle mass.  Therefore, the BDT may learn to reject the background in the training using the fact that the mass of the background and signal candidates is different.  This is just an artifact of how the background sample is obtained and will not be true for background candidates {\em under} the signal peak.  Figure~\ref{fig:md_results} shows the background mis-indentifcation rate {\em vs} $D$ candidate mass.  AdaBoost has clearly learned to use this mis-match in signal and background candidate masses in the training.   The background in the region of the signal is about three times higher than one would expect from looking only at the sideband data.

\begin{figure}[] 
  \centering 
  \includegraphics[width=0.49\textwidth]{mD_train.pdf}
  \includegraphics[width=0.49\textwidth]{MD_eff.pdf}
  \caption{\label{fig:md_results} (left) Signal and (filled) background samples used in training.  (right) Background mis-identification {\em vs} $D$ candidate mass for an arbitrary working point of 10\% background mis-identification in the training region $1.75 < m(D) < 1.85$~GeV is displayed.}
\end{figure}

Figure~\ref{fig:md_results} also shows the background mis-indentifcation rate {\em vs} $D$ candidate mass for the various uniform classifiers where $y = m(D)$ and the choice is for uniformity in the background efficiency\footnote{The algorithms in this paper can easily be made uniform on signal, background or both.}.  
The uBoost algorithm does better than AdaBoost here but is still not optimal.  The way that uBoost achieves uniformity is not such that it can be trusted to work outside the region of training.  The algorithms presented in this paper each does well in achieving similar performance in the training and signal regions.  
Consider, {\em e.g.}, the UGBFL approach to achieving uniform selection efficiency.  In this case the training drives the BDT response itself to have the same PDF everywhere in the region $1.75 < m(D) < 1.85$~GeV (the training region).  
This does not gaurantee that the BDT response is truly independent of $m(D)$ outside the training region, but does strongly suppress learning to use $m(D)$ and in this example results in the desired behavior.  
Finally, if both high and low $m(D)$ sidebands had been used, it is possible for a BDT to create a fake peak near the signal peak location.  The use of UGBFL greatly reduces the chances and possible size of such an effect.

### 2014-10-04 assets/paperdraft/parts/timings.tex

The main drawback of uBoost technique is it's high computational complexity: 
while simple AdaBoost trains $M$ trees, uBoost builds $100 \times M$ trees (contribution of other operations usually can be neglected).

Presented in this paper classifiers are building $M$ trees, though there is more complicated boosting, it takes at each iteration ($k$ is number of neighbors, $N$ is number of events in training sample)

\begin{itemize}
	\item meanAdaBoost: $O(k \times N)$
	\item knnAdaLoss: $O(k \times N)$, for the arbitrary matrix $A$ it is 
	$O( \text{\#nonzero elements in the matrix})$
	\item FlatnessLoss on bins: $O(N \ln N)$
	\item FlatnessLoss on knn: $O(N \ln N + N k \ln k) $ 
\end{itemize}

>>>>

One drawback of the uBoost technique is that it has a high degree of computational complexity: 
while AdaBoost trains $M$ trees (a user-defined number), uBoost builds $100 \times M$ trees.  
The algorithms presented in this paper only build $M$ trees; however, the boosting involves some more complicated algorithms.  Training each of the $M$ trees scales as follows for $N$ training events:
\begin{itemize}
       \item kNNAdaBoost: $O(k \times N)$;
	\item UGBkNNknn: $O(k \times N)$ for $A_{\rm knn}$, and  
	$O( \text{\#nonzero elements in the matrix})$ for arbitrary matrix $A$;
	\item UGBFL(bin): $O(N \ln N)$;
	\item UGBFL(kNN): $O(N \ln N + N k \ln k) $.
\end{itemize}
In the example analysis studied in this paper, we find that the training time for these new algorithms is within a factor of two the same as AdaBoost.  The CPU-resource usage of these new algorithms is not prohibative.
